Renewal and stability in populations structured by remaining years of life

### 2015-10-19 ThanoRepro/ThanoRepro.tex
added force of increment paper

\onehalfspacing
\section*{Introduction}
In this paper, I explore some formal demographic consequences of a particular redefinition of age. Instead of counting age as the time passed since birth, consider age as the amount of time left until death\footnote{This quantity is referred to as \textit{residual} lifetime
in the reliability literaure, or just time-to-death (TTD) elsewhere.}. The first
kind of age is called chronological age, and the second kind of age is called
thanatological age. Individuals in this case move in the same direction along
imaginary life lines, but the reference point for thanatological age is at
the end of the life line instead of at the beginning. For individuals, the timing of events in life is exactly identified using either of the two definitions of age, but
aggregate patterns differ. Demography is built upon a small set of empirical
regularities with respect to chronological age patterns. The models and methods
of demography are in varying degrees prefaced on such regularity. There is evidence that many demographic phenomena are best described in the
aggregatate as a function of time since birth, and others of time until death,
while still others can be a function of both to some degree
\citep{riffe2015ttd}. This observation ought to motivate the formal demographic
base to incorporate this aspect of time in into models. 
%Temporal perspectives and rescalings of the lifecourse are abundant in
% demography already. Such work has enriched the discipline by producing new insights, questions, and conundrums.
% Fertility and reproduction are plausibly a function of both age perspectives,
 % and the question of how to best partition them over time is not investigated here.

I describe the Lotka-Euler-Leslie renewal model when structured by
thanatological age. The renewal model relates fertility, mortality, and population structure in a parsimonious form, revealing the long term consequences of a particular set of conditions. All diverse lines
of contemporary demographic research are relatable to the renewal model, and by
extention to stable population theory. Some might say that stable population
theory is not relevant to everyday demography, since populations are always changing. Even if change precludes the
notion of stability as a prediction, models of long-run consequnces still give a
useful measure of the demographic force implied by a given set of rates.
In this and other ways, stability is never far away. That population projections
lead into a stable state means that some form of stability is implied by the routine practice of demography.
Stability or even stationarity are imaginable for the world population of humans
one day. Models of stationary populations are easily applied by analogy to
closed or nearly-closed birth cohorts. Finally, stable population theory is
relavant for modelling some non-human populations.

%We have stated that many lifecourse transitions covary more with remaining
%lifetime than with life lived. In an even more general sense, by thinking about
% remaining lifetime as a particular class of time-to-transition, one may generalize the lifetable methods presented here into new area applications, such as education, partnering, the economic lifecycle, or migration.

To present the renewal model based on thanatological age is not necessarily a
suggestion for a better way to model or project population. Rather, the
thanatological renewal model represents an alternative and coherent way to
describe the process of population renewal by relating renewal to a death flow
rather than to a birth flow. The chronological-age classified renewal model and
the thanatological renewal model both describe the movement of population stocks
over time, but from opposite vantage points. Specifying a model of thanatological renewal expands the toolbox available to approach complex
temporal population structure. The next section of this article describes the lifetable method of transforming age-classified
data into thanatological age classes. I then derive the continuous version of
the thanatological renewal model, and some other stable properties. This is
followed by the matrix version of the model, which is explained alongside the
lifecycle graph.

\section*{A lifetable approximation of thanatological age}
Direct information on the timing of demographic transitions with respect to
death is available retrospectively from linked population registers or
panel surveys with mortality follow-ups.
Some surveys also ask respondents in various ways about their perceived future lifetime, which may underly those
demographic transitions governed to some degree by volition. The present inquiry
draws only from lifetable-projected remaining lifetime, which
is intended to approximate actual remaining lifetime under the strong
assumption of homogenous populations. Chronological age-classified counts are
redistributed over future death times according to a fixed mortality schedule.
This allows us to make statements about the remaining lifetime distribution
within a stock of population, under certain assumptions.

A lifetable approach to transforming chronologically classified data into
thanatologically classified data is valid when the given lifetable accurately
describes the future mortality of age-classified population. To make
a projective statement about a current stock of population, one does well to
factor in future changes in mortality, and to allow the trajectory of changes
to vary over ages and cohorts. In the case of a stable population model,
mortality is assumed invariant over time, which means that the same survival
pattern underlies each birth cohort. This is the case we explore, and so the
following formulas assume a single lifetable pattern.

The basic transformation to thanatological age proceeds by defining a
deaths density function, similar the lifetable $d(a)$, but over two
dimensions: chronological age attained, $a$, and remaining years of life, $y$.
This function we denote $f(y|a)$ --- The density distribution of thanatological
age conditional on chronological age. This is essentially the same notion as the
familiar $d(a)$, but rescaled as a function of chronological age. For ages higher than 0, simply condition
on survival:\footnote{In the reliability literature, our concept of $d(y|a)$ is denoted by \begin{equation} X - t | X > t
\end{equation}
where $X$ is the total lifespan and $t$ is the current chronological age
attained.}
\begin{equation}
\label{eq:vaupel1}
f(y | a) = \mu(a+y)\frac{\ell(a+y)}{\ell(a)} \ep
\end{equation}

A convenient discretization of \eqref{eq:vaupel1} is to simply work with the
$d_a$ column of the lifetable (here with single ages):
\begin{equation}
\label{eq:dyadiscret}
f_{y, a} =  \frac{d_{a+y}+d_{a+y+1}}{L_a + L_{a+1}} \ec
\end{equation}
where $L_a$ is lifetable exposure.\footnote{Equation~\eqref{eq:dyadiscret}
will incur some error depending on the method used to derive $L_a$. To remove
this error, rescale over the $a$ margin so that each distribution conditioned on
$a$ sums to 1. Working with $\ell_a$ rather than $L_a$ eliminates error and
yields similar results, but is not necessarily better.} One produces in this way a probability density function over future death times for each chronological age,
$a$. The $f(y|a)$- weighted average of $y$ for a given $a$ returns the familiar
remaining life expectancy column of the lifetable, $e(a)$. If $P(a)$ is a
chronological age-classified population count, we can derive $P(y)$, a
thanatological age-classified population count using

@@

\citet{brouard1986structure} applied this method to highlight regional
differences in French age structure, at times combining this projective method
with historical stocks. \citet{brouard1989mouvements}, and later
\citet{vaupel2009life}, showed that in stationary populations the chronological
and thanatological age structures are identical, and this provides a
limited basis for comparison between the two perspectives. \citet{wachter2014essential} hints at, and
\citet{vaupel2014stable} elaborates on how to make chronological and
thanatological age structure commensurate under a constant growth rate, $r$.
These results will also follow from the following specification of a
thanatological renewal model.

@@

\begin{enumerate}
  \item Derive a first rough estimate of the mean remaining years of life at
  reproduction, $\widehat{T^\star}$, akin to Lotka's mean generation time, $T$.
  To start, a good-enough guess is to just assume $r=0$:
\begin{equation}
\widehat{T^\star} = \frac{\int _{y=0}^\infty \int _{a=0}^\infty y d (a+y)
\gamma(y) \dd a \dd y}{\int _{y=0}^\infty \int _{a=0}^\infty d(a+y) \gamma(y)
\dd a \dd y}
\end{equation}
  \item A first rough guess at the net reproduction rate, $R_0$ is given by
 \begin{equation}
  R_0 = \int _{y=0}^\infty \int _{a=0}^\infty d(a+y) \gamma(y) \dd a
\dd y
\end{equation}
  \item A first rough estimate of $r$, $r^0$, is given by
   \begin{equation}
   r^0 = \frac{ln(R_0)}{\widehat{T^\star}}
   \end{equation}
  \item Plug $r^0$ into Equation~\eqref{eq:thanoren} to calculate a
  residual, $\delta^0$.
  \item Use $\delta^0$ and $\widehat{T^\star}$ to calibrate the estimate of $r$
  using
  \begin{equation}
  r^{1} = r^0 + \frac{\delta^0}{\widehat{T^\star} - \frac{\delta^0}{r^0}}
  \end{equation}
  \item Repeat step (4) to to derive a new $\delta^i$, then step (5) to refine
  $r^i$, until converging on a stable $r$ after some 20 or so iterations,
  depending on the degree of precision desired ($\widehat{T^\star}$ is not updated
  in this process).
\end{enumerate}

>>>>

\onehalfspacing
\section*{Introduction}
In this paper, I explore some formal demographic consequences of a particular redefinition of age. Instead of counting age as the time passed since birth, consider age as the amount of time left until death\footnote{This quantity is referred to as \textit{residual} lifetime
in the reliability literaure, or just time-to-death (TTD) elsewhere.}. The first
kind of age is called chronological age, and the second kind of age is called
thanatological age. Individuals in this case move in the same direction along
imaginary life lines, but the reference point for thanatological age is at
the end of the life line instead of at the beginning. For individuals, the timing of events in life is exactly identified using either of the two definitions of age, but
aggregate patterns differ. Demography is built upon a small set of empirical
regularities with respect to chronological age patterns. The models and methods
of demography are in varying degrees prefaced on such regularity. There is evidence that many demographic phenomena are best described in the
aggregatate as a function of time since birth, and others of time until death,
while others can be a function of both to some degree
\citep{riffe2015ttd}. This observation ought to motivate the formal demographic
base to incorporate this aspect of time in into models. 
%Temporal perspectives and rescalings of the lifecourse are abundant in
% demography already. Such work has enriched the discipline by producing new insights, questions, and conundrums.
% Fertility and reproduction are plausibly a function of both age perspectives,
 % and the question of how to best partition them over time is not investigated here.

I describe the Lotka-Euler-Leslie renewal model when structured by
thanatological age. The renewal model relates
fertility, mortality, and population structure in a parsimonious form, revealing the long term consequences of a particular set of conditions. All diverse lines of contemporary demographic research are relatable to the renewal model, and by extention to stable population theory.
%Some might say that stable population
%theory is not relevant to everyday demography, since populations are always
% changing. Even if change precludes the notion of stability as a prediction, models of long-run consequnces still give a
%useful measure of the demographic force implied by a given set of rates.
%In this and other ways, stability is never far away. That population
% projections lead into a stable state means that some form of stability is implied by the routine practice of demography.
%Stability or even stationarity are imaginable for the world population of
% humans one day. Models of stationary populations are easily applied by analogy to
%closed or nearly-closed birth cohorts. Finally, stable population theory is
%relavant for modelling some non-human populations. 
%We have stated that many lifecourse transitions covary more with remaining
%lifetime than with life lived. In an even more general sense, by thinking about
% remaining lifetime as a particular class of time-to-transition, one may generalize the lifetable methods presented here into new area applications, such as education, partnering, the economic lifecycle, or migration.
To present the renewal model based on thanatological age is not necessarily a
suggestion for a better way to model or project population. Rather, the
thanatological renewal model represents an alternative and coherent way to
describe the process of population renewal by relating renewal to a death flow
rather than to a birth flow. I have described a limited case of this
perspective on renewal for the case of stationary populations
\citep{riffe2015force}, and here aim to provide a broader treatment allowing for
growth in the population.
The chronological-age classified renewal model and the thanatological renewal model both describe the movement of population stocks over time, but from opposite vantage points. Specifying a model of thanatological renewal expands the toolbox available to approach complex
temporal population structure. The next section of this article describes the lifetable method of transforming age-classified
data into thanatological age classes. I then derive the continuous version of
the thanatological renewal model, and some other stable properties. This is
followed by the matrix version of the model, which is explained alongside the
lifecycle graph.

\section*{A lifetable approximation of thanatological age}
%Direct information on the timing of demographic transitions with respect to
%death is available retrospectively from linked population registers or
%panel surveys with mortality follow-ups.
%Some surveys also ask respondents in various ways about their perceived future
% lifetime, which may underly those demographic transitions governed to some degree by volition. 
In general, to make a lifetable-based projective statement about the remaining
time distribution of a current stock of population, it is best to to account for population subgroups
with different mortality and to account for uncertainty about future changes in
mortality.
The present inquiry approximates thanatological age using lifetable-projected remaining lifetime,
which is intended to approximate remaining lifetime under the strong
assumptions of homogenous populations und fixed vital rates. 
%Chronological age-classified counts are redistributed over future death times
% according to a fixed mortality schedule.
%In general, a lifetable approach to transforming chronologically classified
%data into thanatologically classified data is sound if the given lifetable
%accurately describes the future mortality of the age-classified population, and
%if the chronological and thantological age patterns are independent.
In this case, each birth cohort undergoes the same
pattern of attrition. In a period setting, this means that we can distribute the
remaining lifespans of the population in each age class using the
age-conditioned lifetable.
%This is the case we explore, and so the following formulas assume a single lifetable pattern.

The basic transformation to thanatological age proceeds by defining a
conditional deaths density function, $f(y|a)$. For chronological
age zero, $f(y|0)$ is identical to $d(a)$. For ages higher than 0, simply
condition on survival:\footnote{In the reliability literature, $f(y|a)$ is
denoted by $ X - t | X > t $, where $X$ is the total lifespan and $t$ is the
current chronological age attained.}
\begin{equation}
\label{eq:vaupel1}
f(y | a) = \mu(a+y)\frac{\ell(a+y)}{\ell(a)} \ep
\end{equation}

A convenient discretization of \eqref{eq:vaupel1} is to simply work with the
$d_a$ column of the lifetable (here with single ages):
\begin{equation}
\label{eq:dyadiscret}
f_{y, a} =  \frac{d_{a+y}+d_{a+y+1}}{L_a + L_{a+1}} \ec
\end{equation}
where $L_a$ is lifetable exposure.\footnote{Equation~\eqref{eq:dyadiscret}
will incur some error depending on the method used to derive $L_a$. To remove
this error, rescale over the $a$ margin so that each distribution conditioned on
$a$ sums to 1. Working with $\ell_a$ rather than $L_a$ eliminates error and
yields similar results, but is not necessarily better.} One produces in this way a probability density function over future death times for each chronological age,
$a$. The $f(y|a)$- weighted average of $y$ for a given $a$ returns the familiar
remaining life expectancy column of the lifetable, $e(a)$. If $P(a)$ is a
chronological age-classified population count, we can derive $P(y)$, a
thanatological age-classified population count using\footnote{An identical
definition is found in \citet{brouard1989mouvements}.}

@@

\citet{brouard1986structure} applied this method to highlight regional
differences in French age structure, at times combining this projective method
with historical stocks. \citet{brouard1989mouvements}, and later
\citet{vaupel2009life}, showed that in stationary populations the
marginal chronological and thanatological age structures are identical, and this
provides a limited basis for comparison between the two perspectives. 
%\citet{wachter2014essential} hints at, and
%\citet{vaupel2014stable} elaborates on how to make chronological and
%thanatological age structure commensurate under a constant growth rate, $r$.
These results will also follow from the following specification of a
thanatological renewal model.

@@

\begin{enumerate}
  \item Derive a first rough estimate of the mean remaining years of life at
  reproduction, $\widehat{T^\star}$, akin to Lotka's mean generation time, $T$.
  To start, a good-enough guess is to just assume $r=0$:\footnote{Using the
  Brouard-Carey equality, the denominator could also be: $\int _{y=0}^\infty
  \ell(y) \gamma(y) \dd y$}
\begin{equation}
\widehat{T^\star} = \frac{\int _{y=0}^\infty \int _{a=0}^\infty y d (a+y)
\gamma(y) \dd a \dd y}{\int _{y=0}^\infty \int _{a=0}^\infty d(a+y) \gamma(y)
\dd a \dd y}
\end{equation}
  \item A first rough guess at the net reproduction rate, $R_0^\star$ is given
  by
 \begin{equation}
  R_0^\star = \int _{y=0}^\infty \int _{a=0}^\infty d(a+y) \gamma(y) \dd a
\dd y
\end{equation}
  \item A first rough estimate of $r$, $r^0$, is given by
   \begin{equation}
   r^0 = \frac{ln(R_0^\star)}{\widehat{T^\star}}
   \end{equation}
  \item Plug $r^0$ into Equation~\eqref{eq:thanoren} to calculate a
  residual, $\delta^0$.
  \item Use $\delta^0$ and $\widehat{T^\star}$ to calibrate the estimate of $r$
  using
  \begin{equation}
  r^{1} = r^0 + \frac{\delta^0}{\widehat{T^\star} - \frac{\delta^0}{r^0}}
  \end{equation}
  \item Repeat step (4) to to derive a new $\delta^i$, then step (5) to refine
  $r^i$, until converging on a stable $r$ after some 20 or so iterations,
  depending on the degree of precision desired ($\widehat{T^\star}$ is not updated
  in this process).
\end{enumerate}
